---
title: "ETC5521 Assignment 1"
subtitle: "Beach Volleyball Stats"
team: [emu]
author:
  - [Justin Thomas]
  - [Mayunk Bharadwaj]
date: "`r Sys.Date()`"
output: 
  html_document
---

[This assignment is for ETC5521 Assignment 1 by Team `r rmarkdown::metadata$team` comprising of `r knitr::combine_words(rmarkdown::metadata$author)`.]{style="color:#006DAE;"}

```{r}
library(tidyverse)
library(readr)
```


# Introduction and motivation

[FILL] Give the bigger picture of the data, and inspire the reader to learn more about the problem by reading your analysis. 

# Data

```{r data-read-in}

vb_matches <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-19/vb_matches.csv', guess_max = 76000)
```

# Data description

The source of our data set comes from a github repo, https://github.com/BigTimeStats/beach-volleyball created by Adam Vagner. This data set provides beach volleyball statistics for men's and women's matches at two major tournaments, the Fédération Internationale de Volleyball (FIVB) Beach Volleyball World Championships and the Association of Volleyball Professionals (AVP) tour. The data provided ranges from September 2000 to August 2019 and it has been collected by the data recorded at the tournaments.

The structure of the data set is:
Rows: 76756
Columns: 65
Data types: Character, Numeric, Data and Difftime  

There are 65 variables in this data set:

  1)"circuit"                 2) "tournament"             3) "country"              
  4) "year"                   5) "date"                   6) "gender"               
  7) "match_num"              8) "w_player1"              9) "w_p1_birthdate"       
 10) "w_p1_age"              11) "w_p1_hgt"              12) "w_p1_country"         
 13) "w_player2"             14) "w_p2_birthdate"        15) "w_p2_age"             
 16) "w_p2_hgt"              17) "w_p2_country"          18) "w_rank"               
 19) "l_player1"             20) "l_p1_birthdate"        21) "l_p1_age"             
 22) "l_p1_hgt"              23) "l_p1_country"          24) "l_player2"            
 25) "l_p2_birthdate"        26) "l_p2_age"              27) "l_p2_hgt"             
 28) "l_p2_country"          29) "l_rank"                30) "score"                
 31) "duration"              32) "bracket"               33) "round"                
 34) "w_p1_tot_attacks"      35) "w_p1_tot_kills"        36) "w_p1_tot_errors"      
 37) "w_p1_tot_hitpct"       38) "w_p1_tot_aces"         39) "w_p1_tot_serve_errors"
 40) "w_p1_tot_blocks"       41) "w_p1_tot_digs"         42) "w_p2_tot_attacks"     
 43) "w_p2_tot_kills"        44) "w_p2_tot_errors"       45) "w_p2_tot_hitpct"      
 46) "w_p2_tot_aces"         47) "w_p2_tot_serve_errors" 48) "w_p2_tot_blocks"      
 49) "w_p2_tot_digs"         50) "l_p1_tot_attacks"      51) "l_p1_tot_kills"       
 52) "l_p1_tot_errors"       53) "l_p1_tot_hitpct"       54) "l_p1_tot_aces"        
 55) "l_p1_tot_serve_errors" 56) "l_p1_tot_blocks"       57) "l_p1_tot_digs"        
 58) "l_p2_tot_attacks"      59) "l_p2_tot_kills"        60) "l_p2_tot_errors"      
 61) "l_p2_tot_hitpct"       62) "l_p2_tot_aces"         63) "l_p2_tot_serve_errors"
 64) "l_p2_tot_blocks"       65) "l_p2_tot_digs"


# Analysis and findings

[FILL] Should include at least one plot or numerical summary for each of your questions, that helps the reader arrive at an answer. You should also write paragraphs describing the methods, summaries and findings. 

# References
